---
title: 'History of LA Redlining'
author: "{Diana Navarro}"
output: 
  html_document:
    df_print: paged
date: "2023-12-15"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = TRUE, message = FALSE, warning = FALSE)
```

## Overview

Present-day environmental justice may reflect legacies of injustice in
the past. The United States has a long history of racial segregation
which is still visible. During the 1930's the Home Owners' Loan
Corporation (HOLC), as part of the New Deal, rated neighborhoods based
on their perceived safety for real estate investment. Their ranking
system, (A (green), B (blue), C (yellow), D (red)) was then used to
block access to loans for home ownership. Colloquially known as
"redlining", this practice has had widely-documented consequences not
only for community wealth, but also health.[^1] Redlined neighborhoods
have less greenery[^2] and are hotter than other neighborhoods.[^3]

[^1]: Gee, G. C. (2008). A multilevel analysis of the relationship
    between institutional and individual racial discrimination and
    health status. American journal of public health, 98(Supplement_1),
    S48-S56.

[^2]: Nardone, A., Rudolph, K. E., Morello-Frosch, R., & Casey, J. A.
    (2021). Redlines and greenspace: the relationship between historical
    redlining and 2010 greenspace across the United States.
    Environmental health perspectives, 129(1), 017006.

[^3]: Hoffman, J. S., Shandas, V., & Pendleton, N. (2020). The effects
    of historical housing policies on resident exposure to intra-urban
    heat: a study of 108 US urban areas. Climate, 8(1), 12.

Check out coverage by the [New York
Times](https://www.nytimes.com/interactive/2020/08/24/climate/racism-redlining-cities-global-warming.html).

A recent study found that redlining has not only affected the
environments communities are exposed to, it has also shaped our
observations of biodiversity.[^4] Community or citizen science, whereby
individuals share observations of species, is generating an enormous
volume of data. Ellis-Soto and co-authors found that redlined
neighborhoods remain the most undersampled areas across 195 US cities.
This gap is highly concerning, because conservation decisions are made
based on these data.

[^4]: Ellis-Soto, D., Chapman, M., & Locke, D. H. (2023). Historical
    redlining is associated with increasing geographical disparities in
    bird biodiversity sampling in the United States. Nature Human
    Behaviour, 1-9.

Check out coverage by
[EOS](https://eos.org/articles/bird-biodiversity-reports-reflect-cities-redlined-past?utm_source=EosBuzz&mkt_tok=OTg3LUlHVC01NzIAAAGOx3ckQPpeDY3QyFKdUnBSYqILj3qLrgb_Q2lasphwcF90xEzMgjZFUo3NHsw3caXDcZe6I8D8qdVsSRr3wzhb6LKAQRgNo8ffhjieEbDwMw).

### Data

#### EJScreen

I will be working with data from the United States Environmental
Protection Agency's EJScreen: Environmental Justice Screening and
Mapping Tool.

According to the [US EPA
website](https://www.epa.gov/ejscreen/purposes-and-uses-ejscreen):

> This screening tool and data may be of interest to community residents
> or other stakeholders as they search for environmental or demographic
> information. It can also support a wide range of research and policy
> goals. The public has used EJScreen in many different locations and in
> many different ways.
>
> EPA is sharing EJScreen with the public:
>
> -   to be more transparent about how we consider environmental justice
>     in our work,
>
> -   to assist our stakeholders in making informed decisions about
>     pursuing environmental justice and,
>
> -   to create a common starting point between the agency and the
>     public when looking at issues related to environmental justice.

EJScreen provides on environmental and demographic information for the
US at the Census [tract](https://en.wikipedia.org/wiki/Census_tract) and
[block group](https://en.wikipedia.org/wiki/Census_block_group) levels.
You will be working with block group data that has been downloaded from
the [EPA site](https://www.epa.gov/ejscreen/download-ejscreen-data). To
understand the associated data columns, you will need to explore the
Technical Documentation and column description spreadsheet available in
the data folder. I also encourage you to explore the [limitations and
caveats](https://www.epa.gov/ejscreen/limitations-and-caveats-using-ejscreen)
of the data.

#### Mapping Inequality

A team of researchers, led by the [Digital Scholarship
Lab](https://dsl.richmond.edu/) at the University of Richmond have
digitized maps and information from the HOLC as part of the [Mapping
Inequality](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58)
project.

I will be working with maps of HOLC grade designations for Los Angeles.
Information on the data can be found
[here](https://dsl.richmond.edu/panorama/redlining/#loc=5/39.1/-94.58&text=downloads).[^5]

[^5]: Robert K. Nelson, LaDale Winling, Richard Marciano, Nathan
    Connolly, et al., "Mapping Inequality," American Panorama, ed.
    Robert K. Nelson and Edward L. Ayers, accessed October 17, 2023,
    <https://dsl.richmond.edu/panorama/redlining/>

#### Biodiversity observations

The [Global Biodiversity Information Facility](gbif.org) is the largest
aggregator of biodiversity observations in the world. Observations
typically include a location and date that a species was observed, and
for this project, I will be working with 2021 data onward.

### Investigate the legacy of redlining in current environmental (in)justice

#### Load the appropriate libraries

```{r }
library(tidyverse)
library(sf)
library(raster)
library(spData)
library(spDataLarge)
library(tidyverse)
library(ggspatial)
library(patchwork)
library(tmap)
```

Reading in EJScreen data and filtering to Los Angeles County.

```{r }
# read in geodatabase of EJScreen data at the Census Block Group level
ejscreen <- st_read("data/EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb/") 

# filter to a county you are interested in
la_county <- ejscreen %>%
  filter(CNTY_NAME %in% c("Los Angeles County"))

#this is for centroid data
county_blockgroup_centroids <- st_centroid(la_county)
```

In this section, I am creating a visualization of:

-   Wastewater discharge by census block groups.

-   Indicating which census block groups are above the 95th percentile
    of national values for wastewater discharge.

    -   Adding a centroid point to display this.

```{r}
#grab only the data above the 95th percentile
centroid95 <- county_blockgroup_centroids %>% 
  filter(P_PWDIS > 95)

#plot the points
tm_shape(la_county) +
  tm_fill(fill = "P_PWDIS", 
          title = "Wastewater discharge", 
          palette = "viridis") +
  tm_shape(centroid95) +
  tm_dots(size = 0.10, 
          aes(color = "red"), 
          fill_alpha = 1.00) +
  tm_add_legend(type="fill",
					  title="Census block groups",
					  labels = c('Above 95 percentile'),
					  fill = "red") +
  tm_grid(crs= 4326, labels.size = .4)+
  tm_scale_bar()+
  tm_compass(position = c("left", "bottom"), 
             size = 1.0) +
  tm_xlab("Longitude")+
  tm_ylab("Latitude") +
  tm_title("Wastewater Discharge by Census Block Group", size = .5) #it cuts off as I make the size larger so had to set to 0.5
```

Now, I am finding the percent of census block groups that:

-   have less than 5% of the population is considered low income

-   are above the 80th percentile for Particulate Matter 2.5

-   are above the 80th percentile for Superfund proximity.

```{r }
low_income <- la_county %>% 
  filter(LOWINCPCT < 0.05)

#nrow(low_income)
nrow(low_income)/ nrow(la_county) * 100

pm25_cbg <- la_county %>% 
  filter(P_PM25 > 80 & P_PNPL > 80)

nrow(pm25_cbg) / nrow(la_county) * 100

```

**The data indicates that the percent of census block groups that have
less than 5% of the population is considered low income is 6.11%. The
percent of census block groups that are above the 80th percentile for
Particulate Matter 2.5 AND above the 80th percentile for Superfund
proximity is 17.35 %.**

Importing redlining information for Los Angeles.

```{r }
LA_redlining <- st_read("https://dsl.richmond.edu/panorama/redlining/static/citiesData/CALosAngeles1939/geojson.json") %>%
  st_make_valid()

```

Visualizing historic redlining boundaries in LA, by HOLC grade:

```{r}
tm_shape(LA_redlining) +
  tm_fill(fill = "grade", 
          title = "HOLC Grade", 
          palette = "magma") +
  tm_shape(la_county) +
  tm_borders(col_alpha = 0.1) +
  tm_graticules() +
  tm_scalebar() + 
  tm_compass(position = c("right", "top"), 
             size = 1)+
  tm_title("Historical redlining boundaries by HOLC grade") +
  tm_xlab("Longitude") +
  tm_ylab("Latitude") 
```

Now I want to find the number of census block groups that fall within
the areas with HOLC grades.

```{r eval = TRUE}
#check out the crs of each 
#st_crs(la_county) == st_crs(LA_redlining)
#change the crs using st transform
actual_la <- st_transform(la_county, crs = 4326)

#length of st_intersect would be the number of census blocks
#crs_match <- st_intersection(LA_redlining,actual_la)

# try using the st_intersects functions

sel_sgbp <- st_intersects(x = actual_la, y = LA_redlining) # returns binary predicate list
sel_logical = lengths(sel_sgbp) > 0 # create logical of which items do intersect
redlining_cbgs2 <- actual_la[sel_logical, ] # filter census block groups based on logicals

# the number of census block groups
print(paste("The number of census block groups that fall within areas with HOLC grads are:", nrow(redlining_cbgs2)))
```

**The number of census block groups that fall within areas with HOLC
grads are 3991.**

In this next part, I wanted to summarize current conditions based on
EJScreen data within historical redlining categories using the mean of
the following variables:

-   \% low income.

-   percentile for particulate Matter 2.5

-   percentile for low life expectancy.

-   percentile for air toxics cancer risk

```{r }
#create data set
la_county1 <- st_join(LA_redlining, actual_la, join = st_intersects)

#check for NAs in datasets
#sum(is.na(la_county1$P_LIFEEXPPCT))
#sum(is.na(la_county1$P_CANCER))

#mean of each cols by holc_grade
mean_values <- la_county1 %>%
  group_by(grade) %>%
  summarize(
    lowinc_mean = mean(LOWINCPCT[grade == "A" | 
                                  grade == "B" | 
                                  grade == "C" | 
                                  grade == "D"], na.rm = TRUE),
    PM25_mean = mean(P_PM25[grade == "A" | 
                                  grade == "B" | 
                                  grade == "C" | 
                                  grade == "D"], na.rm = TRUE),
    lifeexp_mean = mean(P_LIFEEXPPCT[grade == "A" | 
                                  grade == "B" | 
                                  grade == "C" | 
                                  grade == "D"], na.rm = TRUE),
    cancerisk_mean = mean(P_CANCER[grade == "A" | 
                                  grade == "B" | 
                                  grade == "C" | 
                                  grade == "D"], na.rm = TRUE)
    )

print(mean_values) 

```

**These means show that HOLC grad D tends to have higher means on
average than A, B, and C. If you take a look at the map of Los Angeles
County, you could see that the inner city is mostly HOLC grad D, and
falls in this order from higher density to lower: D, C, B, and A. This
means that on average they have lower income percentage, and are in the
area with higher percentile for particulate matter 2.5, they are in the
higher percentile for low life expectancy, and higher percentile for air
toxics cancer risk.**

### Investigate the legacy of redlining in biodiversity observations

In order to understand how many bird observations fall within
neighborhoods with HOLC grads, I plotted the distribution of percent of
observations within each redlining categories.

```{r}
#call the data from the file
birdy_data <- st_read("data/gbif-birds-LA") 

#select data from 2022
birdy_year <- birdy_data[birdy_data$year == 2022, ]

#check to make sure the crs lines up 
#st_crs(LA_redlining) == st_crs(birdy_data)

#combine the dataset
birdy_dat1 <- st_join(LA_redlining,birdy_year, join = st_intersects)

#find the percentage of holcgrades
holc_percent <- birdy_dat1 %>%
  st_drop_geometry() %>% 
  group_by(grade) %>%
  summarize(percentage = n()/nrow(birdy_dat1) *100) 

print(holc_percent)

#plot the percentages
ggplot(holc_percent, aes(x = grade, y = percentage)) +
  geom_col(fill = "cornflowerblue")+
  labs(title = "Bird Observations within redlining boundaries in 2022 ", x = "HOLC Grade", y = "Percentage of Bird observations")

```

**Given the historical context, I expected A and B HOLC grads to have
more bird observations, but since this is more recent (only 2022
observations) there has been conservation efforts to conduct more bird
observations within C and D HOLC grads. One reason for the increased
percentage in C and potentially D is gentrification, although on average
they are still 'cold spots' within bird biodiversity observations, it is
possible that only pulling 2022 data can show the effect of
gentrification on bird observations.**
